---
title: "geodimension: Definition of Geographic Dimensions"
author: "Jose Samos (jsamos@ugr.es)"
date: "2020-11-26"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{geodimension: Definition of Geographic Dimensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The *multidimensional data model* was defined with the aim of supporting data analysis. In multidimensional systems, data is structured in facts and dimensions[^1]. The *geographic dimension* plays a fundamental role in multidimensional systems. Apart from the analysis possibilities it offers, like any other dimension, it is very interesting to have the possibility of representing the reports obtained from multidimensional systems, using their geographic dimensions, on a map, or performing spatial analysis on them. This functionality is supported by package [`geomultistar`](https://CRAN.R-project.org/package=geomultistar).

[^1]: Basic concepts of dimensional modelling and star schemas are presented in [`starschemar`](https://CRAN.R-project.org/package=starschemar) vignettes.

To define a geographic dimension in a star schema, we need a table with attributes corresponding to the levels of the dimension. Additionally, we will also need one or more geographic layers to represent the data using this dimension. We can obtain this data from available vector layers of geographic information. In simple cases, one layer is enough; but we often need several layers related to each other. The relationships can be defined by common attribute values or can be inferred from the respective geographic information.

The main objective of this package is to support the definition of geographic dimensions from layers of geographic information that can be used in multidimensional systems. In particular, through package [`geomultistar`](https://cran.r-project.org/package=geomultistar) they can be used directly.

The rest of this document is structured as follows: First, an illustrative example of how the package works is developed. Then, the document ends with conclusions.


# An illustrative example

Suppose we have a multidimensional design on US data and the geographic dimension is defined at the *city* level. For each city we have its name and the code of the state in which it is located. It would be interesting to have other levels of detail in this dimension to be able to perform roll-up operations.

With this objective, we look for layers of geographic information. In [Uited States Census Bureau](https://www.census.gov) we find layers at various levels of detail, including *place*, *county*, *state*, *region*, *division* and *nation*. We have obtained them (for *place* we have selected only the cities) and we have stored[^2] them in `layer_us_city`, `layer_us_county`, `layer_us_state`, `layer_us_region`, `layer_us_division` and `layer_us_nation`, respectively.

[^2]: To take up less space in the package, its geometry has also been simplified.

The layers are related to each other. In some cases they have attributes in common, in others, although there is a relationship, it is not explicitly defined. We can use `geodimension` to support the definition of these relationships. Once defined, it will also offer us support to exploit them and obtain information from them.

Thus, three phases can be distinguished:

- Definition of levels.

- Definition of relationships.

- Obtaining information.

## Definition of levels

Each of the layers that we have obtained represents a conceptual level of the geographical dimension, in the package it is called `geolevel`. To define a `geolevel`, we need a layer and the set of attributes that make up the layer's key (which uniquely identify each of its instances).

We can previously check if a set of attributes form a key of the layer using the `check_key` function.

```{r message=FALSE}
library(tidyr)
library(sf)
library(geodimension)

names(layer_us_city)

check_key(layer_us_city, key = c("name", "statefp"))

check_key(layer_us_city, key = c("geoid"))
```

We might expect the city name (`name`) and state code (`statefp`) to be sufficient to identify a city, however, they are not[^3]. We check that the `geoid` field is a valid key, therefore, it will be the one we use to define the `geolevel`.

[^3]: For our example, this is not a problem, just a warning that we must pay attention when we select the cities, to determine if any are repeated.

We can check the geometry that is considered for the definition of the level by means of the `get_geometry` function (it simplifies the types into three: *point*, *line*, and *polygon*). In addition, we give each level a name to be able to refer to it, as shown below.

```{r}
get_geometry(layer_us_city)

city <-
  geolevel(name = "city",
           layer = layer_us_city,
           key = c("geoid"))
```

For `county` it is the same as for `city`, the name and the code of the state do not compose a valid key. In this case, the geometry is polygon. Additionally, a layer with another geometry can be associated using the `add_geometry` function or, in the case of point-type geometry, it can be generated from the polygon layer using the `complete_point_geometry` function, as is done below.

```{r message=FALSE, warning=FALSE}
get_geometry(layer_us_county)

county <-
  geolevel(name = "county",
           layer = layer_us_county,
           key = c("geoid")) %>%
  complete_point_geometry()
```

In the same way, the checks are made and the rest of the levels are defined. Although all the layers are of the polygon type, we generate the point layers only for those that we think may be needed. Below is only the definition of the levels.

```{r message=FALSE, warning=FALSE}
state <-
  geolevel(name = "state",
           layer = layer_us_state,
           key = c("statefp")) %>%
  complete_point_geometry()

region <-
  geolevel(name = "region",
           layer = layer_us_region,
           key = c("regionce"))

division <-
  geolevel(name = "division",
           layer = layer_us_division,
           key = c("divisionce"))

nation <-
  geolevel(name = "nation",
           layer = layer_us_nation,
           key = c("name"))
```

Once the levels are defined, then we will define the dimension and the relationships between the levels.

## Definition of relationships

To define a `geodimension`, we give it a name and start from any `geolevel`. Next, we add the rest of the `geolevels` in any order.

```{r}
gd <-
  geodimension(name = "gd_us",
               level = region) %>%
  add_level(division) %>%
  add_level(state) %>%
  add_level(nation) %>%
  add_level(city) %>%
  add_level(county)
```

Next, we can define the relationships that we want to consider between the levels. In a relationship there are two parts, the *lower level* and the *upper level*. To define the relationships, the following points must be taken into account:

- There are no restrictions on the relationships you define, as long as the relationship can be established.

- It is only necessary to define direct relationships (from them, indirect ones are derived).

- Relationships can be defined using attributes with common values between the levels or through geographic relationships between their instances.

- If we want to reference the upper level through attributes, these must form a valid key for the level (it does not necessarily have to be the key that was indicated when defining it, this is used by default).

- Currently, to define a relationship using geographic properties, the upper level must be of type polygon.

The relationships between *state*, *region*, *division* and *nation* are defined below.

```{r message=FALSE, warning=FALSE}
gd <- gd %>%
  relate_levels(lower_level_name = "state",
                lower_level_attributes = c("division"),
                upper_level_name = "division") %>%
  relate_levels(lower_level_name = "division",
                upper_level_name = "region",
                by_geography = TRUE) %>%
  relate_levels(lower_level_name = "region",
                upper_level_name = "nation")
```

The relationship between *state* and *division* is defined by a *state* attribute that matches the *division* key. 

The relationship between *division* and *region* is defined by geographic relationship because they do not have any attributes in common.

Finally, as *nation* has only one instance, it relates directly to all instances of the indicated level (in this case, *region*), no additional relationship is required.

In addition to these relationships there is a clear relationship between *city* and *state* and also between *county* and *state*. In both cases it can be defined by attributes.

```{r}
gd <- gd %>%
  relate_levels(lower_level_name = "city",
                lower_level_attributes = c("statefp"),
                upper_level_name = "state") %>%
  relate_levels(lower_level_name = "county",
                lower_level_attributes = c("statefp"),
                upper_level_name = "state")
```

The relationship between *city* and *county* is a bit fuzzy (at least for me), it depends on the state, the city and also the county. In any case, if we are interested in establishing a relationship, we could define it between *city* and *state* using the geographical properties of the levels, since county includes polygon-type geometry.

```{r message=FALSE, warning=FALSE}
gd <- gd %>%
  relate_levels(lower_level_name = "city",
                upper_level_name = "county",
                by_geography = TRUE)
```

We can check if all the instances have been related using the following function:

```{r}
nrow(
  gd %>% get_unrelated_instances(lower_level_name = "city",
                                 upper_level_name = "county")
)
```

Since there are no unrelated instances, each city has been linked to the county whose boundaries contain it.

If we did a similar check for each defined relation, we would observe that all the instances of the lower levels have been related to those of the higher levels, except for the following:

```{r}
nrow(
  gd %>% get_unrelated_instances(lower_level_name = "state",
                                 upper_level_name = "division")
)
```

If we look at the content:

```{r, results = "asis", echo = FALSE}
t <-gd %>% get_unrelated_instances(lower_level_name = "state",
                                 upper_level_name = "division")
pander::pandoc.table(t, split.table = Inf)
```

The region or division code does not correspond to any of the existing ones. If we want to relate all the instances of *state* to the *nation* level, we must define the relationship explicitly (there are other solutions but we will not consider them here, in any case they can be generated from this one).

```{r }
gd <- gd %>%
  relate_levels(lower_level_name = "state",
                upper_level_name = "nation")
```

With these operations we have defined a `geodimension`.

## Obtaining information

From it we can obtain the data table to define a dimension in a star schema or the layer or layers associated with that table at the level we need.

```{r}
ld <- gd %>%
  get_level_data(level_name = "division")
names(ld)

ld <- gd %>%
  get_level_data(level_name = "division",
                 inherited = TRUE)
names(ld)

ll <- gd %>%
  get_level_layer(level_name = "division",
                 inherited = TRUE)
names(ll)
```

If we need the data at another level of detail, we can obtain it in a similar way.

```{r}
ld <- gd %>%
  get_level_data(level_name = "state",
                 inherited = TRUE)
names(ld)

ll <- gd %>%
  get_level_layer(level_name = "state",
                  only_key = TRUE)

plot(ll)
```

In addition to these functions, the package offers other support functions to aid in the definition of levels (for example, to determine the key attributes of a layer), to relate instances of levels whose relationship is not immediately established, or to configure the `geodimension` to obtain a customized output.


# Conclusions

